Rationality, Evolution \& the Environment

### 2015-10-05 paper/paper.tex
sec 2 eddits

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a set of
players, $A_i$ a set of acts and $\pi_i$ a payoff function for each player $i \in N$. The
classical notion of rationality from Definition~\ref{def:rationality} also applies to
games. Let $A_{-i} = \bigtimes A_j$ for all $j \neq i$ be the set of all profiles of acts for
all players excepts $i$. If player $i$ has a belief $\mu_i \in \Delta(A_{-i})$ about the
behavior of his opponents, he chooses rationally only his choice $a^*_i$ maximizes expected
utility $\EU(a) = \sum_{a_{-i} \in A_{-i}} \mu_i \myts \pi_i(a_i, a_{-i})$.

We focus here on symmetric games with two players. A two-player game is symmetric if An example
is given in Figure~\ref{coordgame1}. \textcolor{red}{EXPAND}

\begin{itemize}
\item include here: minimax and choice given a belief
\end{itemize}

\todo[inline]{should we define NE, ESS/NSS here? dominance? we should include \emph{best
    reply}, because it is used in the definition of regret.}

@@

Regret theory has been introduced in decision theory by Savage [] and Niherens [], and later
developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982] independently. Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can solve
game theoretical puzzles (like the Centipede and the Traveller's dilemma) in a way that is
closer to everyday intuition and empirical data. In this paper the notion of regret is defined
as in \citet{HalpernPass2012:Iterated-Regret}:

\begin{definition}[Regret, regret minimization] \label{defn:regret} Given a simultaneous-move
  game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the regret of player $i$ at profile
  $(a_i,a_{-i})$ is $\text{reg}(a_i,a_{-i}):= \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $, where
  $a_i^\$ \in A_i$ denotes $i$'s best reply to action $a_{-i} $. The regret of an action
  $a_i \in A_i$ is then
  $\text{reg}(a_i):= \text{max}_{a_{-i}\in A_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $. An
  action $a^{r}_i $ is regret minimizing if
  $a^{r}_i \in \text{min}_{a_i} \text{max}_{a_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $.
\end{definition}

\begin{example}[Coordination Game]
  Let us consider the coordination game in Figure \ref{coordgame1}. Figure~\ref{coordgame1reg}
  shows the pairwise regrets and the regrets of each row-player action. It is easy to see that
  $\text{reg}(I)=2$: when the other player chooses action $I$ the pairwise regret $reg(I,I)$ is
  $0$ since $I$ is a best reply to $I$, and when the other player chooses $II$ the pairwise
  regret $reg(I, II)$ is $2$ since $\pi_i(II,II)-\pi_i(I,II)=2$. With the same reasoning,
  $\text{reg}(II)=1$. Two regret minimizing players would then choose action $II$, each getting
  a payoff of $2$. In this case regret minimization selects the Pareto-dominant Nash
  equilibrium.\todo{explain Pareto-dominant NE; otherwise skip it?} $ \medsquare $

@@

\begin{example}[The Traveller's Dilemma \citep{HalpernPass2012:Iterated-Regret}]
Let us now consider the Traveller's Dilemma case. The story behind the game goes as
follows. There are two travellers, Ann and Bob, who lost their luggages. Ann and Bob had
exactly the same luggage and they had insured the luggages with the same insurance company. The
insurance policy is that the two travellers have to separately claim a certain amount between
2\$ and 100\$ for the reimbursement, and if they both claim the same amount then they will be
reimbursed by that amount. Otherwise, if one of them claims more than the other, then the one
who claimed the higher amount will get the lower amount minus 2\$, while the one who claimed
the lower amount will get what (s)he claimed plus 2\$. The game is symmetric, and the payoff
function for both players is:
\begin{align*}
  \pi_i(a_i,a_{-i}) = \begin{cases} a_i \text{ \hspace{1cm} if } a_i = a_{-i} \\ a_i + 2 \text{
      \hspace{.4cm} if } a_i < a_{-i} \\ a_{-i} -2 \text{ \hspace{.25cm} if } a_{-i} <
    a_i \end{cases}
\end{align*}
By a simple computation we have:  $\forall a_i \in \lbrace 96\$, ..., 100\$ \rbrace \text{, } \text{reg}(a_i)=3 $ and $\forall a_i \in \lbrace 2\$, ..., 95\$ \rbrace \text{, } \text{reg}(a_i)>3$. If we now eliminate all the actions that do not minimize regret in the first place and iterate the regret minimization only on the action set $\lbrace 96\$, ..., 100\$ \rbrace$, we get that 97\$ is the action that minimize regret. Indeed, $\text{reg}(97)=2$ and $\forall a_i \in \lbrace 96\$, 98\$, 99\$, 100\$ \rbrace \text{, } \text{reg}(a_i)=3 $. Hence, one step of regret minimizing reasoning eliminates all the actions smaller than 96\$, and if we iterate the process once more the unique outcome is 97\$. Finally, it is worth noticing that the only rationalizable equilibrium of the game corresponds to the outcome $(2,2)$, that is consequently the only Nash equilibrium, the only perfect equilibrium, and the only sequential equilibrium of the game. $ \medsquare $

@@

\todo[inline]{add results about regret: reg min is security choice on negative regret transform
(related to Figure 1b); regret choice equal to veridical in case of probabilistic beliefs and
maximization of expected gains}

>>>>

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a set of
players, $A_i$ a set of acts and $\pi_i$ a payoff function for each player $i \in N$. For the
most part, this paper will focus on symmetric games with two players. A two-player game is
symmetric if players have the same action set, $A_1 = A_2$, and payoffs are symmetric:
$\pi_1(\myact{k}, \myact{l}) = \pi_2(\myact{l}, \myact{k})$ for all $k$ and $j$. An example is
given in Figure~\ref{coordgame1}, which shows a coordination game. The numbers give the payoff
for the player choosing the row against an opponent who chooses the column acts. Since the game
is symmetric, it suffices to give just these numbers.

With some due simplification, there are two types of approaches to solving games: one focuses
on the individual agents' perpectives and gives \emph{choice rules}, another considers sets of
choices of all agents that are, in some sense or other, locked-in. Let's look at either in
turn, to the extent necessary for this paper:

\paragraph{Choice rules.} The classical notion of rationality from
Definition~\ref{def:rationality} also applies to games. Let $A_{-i} = \bigtimes A_j$ for all
$j \neq i$ be the set of all profiles of acts for all players except $i$. If player $i$ has a
belief $\mu_i \in \Delta(A_{-i})$ about the behavior of his opponents, he chooses rationally
only if his choice $a^*_i$ maximizes expected utility
$\EU(a) = \sum_{\act_{-i} \in \Acts_{-i}} \mu_i(\act_{-i}) \myts \pi_i(a_i, a_{-i})$. So this
is a choice rule, based on \emph{maximization of expected utilities}. It requires a
probabilistic belief about the opponents' action choices. In the example from
Figure~\ref{coordgame1}, a rational player would choose $I$ exactly when he believes that his
opponent will choose $I$ with a probability at least $\nicefrac{2}{3}$.

If an agent does not have a concrete probabilistic belief (see also
Section~\ref{sec:impr-prob-beli} below), other rules of choice are available. Perhaps the most
well-known, the \emph{maximin rule}, would tell the agent to choose act $\act^*$ only if it
maximizes payoffs for all worst case scenarios, i.e.,
$\act^* \in \argmax_{\myact{i} \in \Acts_i} \min_{\myact{j} \in \Acts_j} \pi_i(\myact{i},
\myact{j})$. For the example in Figure~\ref{coordgame1}, the maximim rule looks at the minima
in each row, asking: what is the worst possible out come for each possible act? It is $0$ in
either case. Consequently, the maximin rule would not favor either act; players choosing by the
maximin rule would be indifferent between acts $I$ and $II$.

Another well-known choice rule is \emph{Laplace's rule} which prescribes to choose $\act^*$ if
$\act^* \in \argmax_{a \in A} \sum_{a'} \pi(a, a')$. Laplace's rule is equivalent to rationally
choosing under a \emph{flat belief} that assigns equal probability to all of the opponent's
choices. In the coordination game of Figure~\ref{coordgame1}, Laplace's rule selects act $II$.

\paragraph{Equilibrium \& stability.}

\begin{itemize}
\item \textcolor{red}{add here:}
  \begin{itemize}
  \item NE, strict, weak
  \item ESS, NSS
  \end{itemize}
\end{itemize}

\todo[inline]{should we also include here: dominance}

@@

Regret theory has been introduced in decision theory by \citet{Savage1951:The-theory-of-s} and
Niherens [], and later developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982]
independently. Lately, \citet{HalpernPass2012:Iterated-Regret} showed how the use of regret
minimization can give solutions to game theoretical puzzles (like the Traveller's dilemma, see
Example~\ref{example:TravelersDilemma}) in a way that is closer to everyday intuition and
empirical data. In this paper the notion of regret is defined as in
\citet{HalpernPass2012:Iterated-Regret}:

\begin{definition}[Regret, regret minimization] \label{defn:regret} Given a simultaneous-move
  game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the \emph{pairwise regret} of player
  $i$ at profile $(a_i,a_{-i})$ is
  $\text{reg}(a_i,a_{-i}):= \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $, where
  $a_i^\$ \in \argmax_{a_i \in A_i} \pi_i(a_i, a_{-i})$ denotes $i$'s best reply to the
  opponents' actions $a_{-i}$. The \emph{regret} of a$a_i$ is then
  $\text{reg}(a_i):= \text{max}_{a_{-i}\in A_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $. An
  action $a^{r}_i $ is regret minimizing if
  $a^{r}_i \in \text{min}_{a_i} \text{max}_{a_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $.
\end{definition}

\begin{example}[Coordination Game]
  Let us consider the coordination game in Figure \ref{coordgame1}. Figure~\ref{coordgame1reg}
  shows the pairwise regrets and the regrets of each row-player action. It is easy to see that
  $\text{reg}(I)=2$: when the other player chooses action $I$ the pairwise regret $reg(I,I)$ is
  $0$ since $I$ is a best reply to $I$, and when the other player chooses $II$ the pairwise
  regret $reg(I, II)$ is $2$ since $\pi_i(II,II)-\pi_i(I,II)=2$. With the same reasoning,
  $\text{reg}(II)=1$. A regret minimizing player would therefore choose action $II$. $ \medsquare $

@@

\begin{example}[The Traveller's Dilemma \citep{HalpernPass2012:Iterated-Regret}]
\label{example:TravelersDilemma}
Let us now consider the Traveller's Dilemma case. The story behind the game goes as
follows. There are two travellers, Ann and Bob, who lost their luggages. Ann and Bob had
exactly the same luggage and they had insured the luggages with the same insurance company. The
insurance policy is that the two travellers have to separately claim a certain amount between
2\$ and 100\$ for the reimbursement, and if they both claim the same amount then they will be
reimbursed by that amount. Otherwise, if one of them claims more than the other, then the one
who claimed the higher amount will get the lower amount minus 2\$, while the one who claimed
the lower amount will get what (s)he claimed plus 2\$. The game is symmetric, and the payoff
function for both players is:
\begin{align*}
  \pi_i(a_i,a_{-i}) = \begin{cases} a_i \text{ \hspace{1cm} if } a_i = a_{-i} \\ a_i + 2 \text{
      \hspace{.4cm} if } a_i < a_{-i} \\ a_{-i} -2 \text{ \hspace{.25cm} if } a_{-i} <
    a_i \end{cases}
\end{align*}
By a simple computation we have:  $\forall a_i \in \lbrace 96\$, ..., 100\$ \rbrace \text{, }
\text{reg}(a_i)=3 $ and $\forall a_i \in \lbrace 2\$, ..., 95\$ \rbrace \text{, }
\text{reg}(a_i)>3$. If we now eliminate all the actions that do not minimize regret in the
first place and iterate the regret minimization only on the action set $\lbrace 96\$, ...,
100\$ \rbrace$, we get that 97\$ is the action that minimize regret. Indeed, $\text{reg}(97)=2$
and $\forall a_i \in \lbrace 96\$, 98\$, 99\$, 100\$ \rbrace \text{, } \text{reg}(a_i)=3
$. Hence, one step of regret minimizing reasoning eliminates all the actions smaller than 96\$,
and if we iterate the process once more the unique outcome is 97\$. Finally, it is worth
noticing that the only rationalizable equilibrium of the game corresponds to the outcome
$(2,2)$, that is consequently the only Nash equilibrium, the only perfect equilibrium, and the
only sequential equilibrium of the game.

\begin{itemize}
\item \textcolor{red}{add here:}
  \begin{itemize}  
  \item outcome of rationalizability (max-EU)
  \item outcome of maximin rule
  \item outcome of Laplace rule
  \end{itemize}

 $ \medsquare $

@@

Regret minimization is intimately related to the maximin rule. Let the \emph{negative regret
  transform} of a strategic game $G$ be the game $G'$ derived from $G$ by substituting the
payoff $\pi_i(a_i, a_{-i})$ of all players in $G$ by the negative pairwise regret
$reg(a_i,a_{-1})$. For example, the negative regret transform of the game in
Figure~\ref{coordgame1} would have the following payoff matrix, i.e., the negatives of the
pairwise regrets given in Figure~\ref{coordgame1}:

\begin{center}
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & -1 & 0 \\
      II & 0 & -2\\
      \bottomrule
    \end{tabular}
\end{center}

\noindent The acts selected by the maximin rule for the negative regret transpose are exactly
the acts selected by regret minimization for the original game.

This means that we can think of the maximin rule and regret minimization as being two
instantiations of the same choice rule, but operating on different subjective representations
of utility: while maximin considers actual objective payoffs, regret minimization considers
non-veridical, regret-based subjective utilities. As the example of Figure~\ref{coordgame1}
shows, the same choice rule can give different behavioral predictions on different preference
representations. While maximin is indifferent between $I$ and $II$, regret minimization
uniquely selects $II$.

When we think of regret as a non-veridical subjective payoff representation, we can also pair
it with other choice rules, such as the maximization of expected utility and Laplace's rule
(which is just a special case of the former). It turns out, however, that this makes no
noticeable difference:

\begin{fact} \label{fact:maxEU-minReg} For any strategic game $G$ and any arbitrary
  probabilistic belief $\mu \in \Delta(A_{-i})$ about the opponents' behavior, the acts
  selected by maximization of expected utility for $G$ under $\mu$ are exactly the acts
  selected by the maximization of expected utility for the negative regret transform of $G$
  under $\mu$ \citep[e.g.][]{HalpernPass2012:Iterated-Regret}.
\end{fact}

\noindent In other words, given the same probabilistic belief, maximization of expected
(objective) payoff and minimization of expected (subjective) regret are behaviorally
equivalent. This holds for arbitrary games, as well as for solitary decision situations.

In sum, we have so far considered four different types of choice rules, each differing along a
preference and a belief dimension. As for preference representations, agents could either
target objective payoffs, or consider non-veridical subjective regret transformations. As for
beliefs, agents could either hold probabilistic belief or not. In the former case, we would
assume that they maximize expectations under whatever preference representation they hold; in
the latter, we would assume that they play maximin based on their preference
representations. There are doubtlessly many more possible preference and belief types, but
these are the main characters in the plot to follow.
